Biogeography of crop progenitors and wild plant resources in the terminal Pleistocene and Early Holocene of West Asia, 14.7–8.3 ka

### 2025-01-10 analysis/appendix_section.qmd
Prepare draft

```{r fig-palaeodist, fig.height=16, fig.cap=paste0("Predicted palaeodistributions of *", taxon, "*")}
if (length(archaeo_pred) > 0) {
  archaeo_geom <- geom_sf(
    aes(shape = .pred_class),
    data = archaeo_pred |>
      mutate(period = climate_period) |>
      distinct(longitude, latitude, period, .keep_all = TRUE) |>
      st_as_sf(coords = c("longitude", "latitude"), crs = latlong)
  )
} else {
  archaeo_geom <- NULL
}

>>>>

```{r fig-palaeodist, warning=FALSE, fig.height=16, fig.cap=paste0("Predicted palaeodistributions of *", taxon, "*")}
if (length(archaeo_pred) > 0) {
  archaeo_geom <- geom_sf(
    aes(shape = .pred_class),
    data = archaeo_pred |>
      mutate(period = climate_period) |>
      distinct(longitude, latitude, period, .keep_all = TRUE) |>
      st_as_sf(coords = c("longitude", "latitude"), crs = latlong)
  )
} else {
  archaeo_geom <- NULL
}

### 2025-01-10 analysis/paper.qmd
Prepare draft

---
title: Biogeography of crop progenitors and wild plant resources in the terminal Pleistocene and Early Holocene of West Asia, 14.7–8.3 ka
date: today
author: 
- name: Joe Roe
  id: jr
  orcid: 0000-0002-1011-1244
  email: joeroe@hey.com
  url: https://joeroe.io
  affiliations:
    - name: University of Bern
      country: Switzerland
    - name: University of Copenhagen
      country: Denmark
  corresponding: true
- name: Amaia Arranz-Otaegui
  id: aao
  orcid: 0000-0002-5091-6426
  affiliation: 
    - name: University of the Basque Country
      country: Spain
abstract: |
  This paper presents the first continuous, spatially-explicit reconstructions 
  of the palaeodistributions of 82<!-- TODO update --> plant species found regularly in association 
  with early agricultural archaeological sites in West Asia, including the 
  progenitors of the first crops. We used machine learning to train an 
  ecological niche model of each species based on its present-day distribution 
  in relation to climate and environmental variables. Predictions of the 
  potential ranges of these species at key stages of the Pleistocene–Holocene 
  transition could then be derived from these models using downsampled data from 
  palaeoclimate simulations.
  The models predict significant reductions and/or shifts in species ranges in
  the terminal Pleistocene and Early Holocene compared to present conditions. 
  Many species that are found throughout the region's 'hilly flanks' today 
  are indicated to have much more restricted distributions centered on the 
  Levant, Cyprus and Western Anatolia. In addition, overall ranges shrunk by an 
  average of c. 25% from the terminal Pleistocene to the Early Holocene. 
  Modelled ranges do not reliably predict the observed  occurrence of specific 
  species at archaeological sites,
  however [...]
  The regional ubiquity of species in the archaeological record is [not] correlated with the 
  predicted size of its range and the diversity of archaeobotanical assemblages is [not] 
  correlated with the predicted diversity of its environs. This indicates that 
  trends in taxonomic composition of the archaeobotanical record is [not] likely 
  to have been influenced by environmental change and species turnover, in
  addition to human economic choices.
format: 
  elsevier-pdf:
    keep-tex: true
    # cite-method: citeproc
    journal:
      name: Open Quaternary
      formatting: preprint
      cite-style: authoryear
bibliography: references.bib
execute: 
  echo: false
---

@@

The Pleistocene–Holocene transition in West Asia marked a turning point in global environmental history, as humans brought the first plants under cultivation and began modifying surrounding ecosystems to support their own subsistence.
West Asia is part of the native range of a remarkable number of domesticable plant species, including wild relatives of wheat, barley, peas, lentils, and other crops of global importance [@cite].
These species supported uniquely dense and complex Late Epipalaeolithic (15–11.7 ka) societies based on intensive foraging [@cite] and eventually pre-domestication cultivation [@cite].
As they were brought under domestication in the Pre-Pottery Neolithic period (11.7–8.5 ka), the world's first agriculture was shaped by the ecosystems from which it emerged and was embedded in.

Decades of research in archaeobotany and zooarchaeology have reconstructed the subsistence economies of Late Epipalaeolithic and Neolithic sites in great detail [@cites].
Together with studies of charcoal [@cite], pollen [@cite], soil [@cite], and a variety of palaeoclimate archives [@JonesEtAl2019], they also tell us much about the environments surrounding these settlements.
However, each of these sources of evidence is subject to the wide variety of taphonomic and recovery biases that are inherent in any direct record of the past.
They are also, by definition, records of the (human) environment at particular times and places.
Interpolating these snapshots to give a holistic picture of the regional ecologies is not straightforward – to date, it has tended to rely on non-explicit, inductive modelling.
The majority are also filtered through human action, producing a mixed single that makes it difficult to disentangle anthropic effects from the background of environmental change in this period of rapid climatic alteration.

@@

# Background

* The transition to agriculture in West Asia was...

@@

## Ecological niche modelling in archaeology

Ecological niche modelling or species distribution modelling is widely used by ecologists to predict the geographic range of a species based on a set of environmental predictors [@FranklinMiller2009].
Essentially, it involves combining records of where an organism has been observed with environmental data (climate, topography, etc.) for those locations to model the range of environmental values at which that species  – its environmental niche.
This model can then be used to predict the range of the organism in question either in the same or a different environment.
@CITE suggests reserving the term 'species distribution modelling' for when the method is used to recover the verifiable range of a species in a real and existing environment, and using 'ecological niche modelling' as the broader term covering hypothetical or predictive applications – a convention we follow here when referring to predictive or 'hindcast' models of past ranges.
Within this overarching framework, ecological niche modelling encompasses a wide range of applications and a variety of potential environmental predictors, modelling approaches, and methodologies, which we will not attempt to review here.

@@

@FranklinEtAl2015 review palaeoecological niche modelling and advocate for its greater adoption in environmental archaeology.
In an early application to West Asia, @ConollyEtAl2012 used the occurrence of wild and domestic *Bos* remains at prehistoric archaeological sites to map the evolving niche of cattle over the Pleistocene–Holocene transition.
It has been used to model the availability of fauna exploited by humans at wider scales [e.g. @deAndresHerreroEtAl2018; @YaworskyEtAl2023] and, in a West Asian context, of foraged plant resources in the landscape around the Neolithic sites on the Konya Plain [@CollinsEtAl2018].
Modelling the spread of crops has been another significant archaeological application [e.g. @CremaEtAl], though not as yet applied to West Asia.

@@

The major practical limitation of the hindcasting approach is that it relies on spatially explicit, high resolution palaeoenvironmental surfaces with continuous coverage of the region and periods of interest.
Until recently, this has not been widely available for most applications, which is perhaps why only a minority of studies use it [cf. @YaworskyEtAl2023].
In this study, we are able to take advantage of the increasing availability of high resolution, global palaeoclimate data derived from simulation experiments with general circulation models of climate [@BrownEtAl2018; @BrownEtAl2020; @KargerEtAl2023].

@@

The aim of our study was to model the biogeography of species relevant to human subsistence economies in West Asia (excluding the Southern Arabian peninsula, see @fig-region) during the archaeological Late Epipalaeolithic (15–11.7 ka) and Pre-Pottery Neolithic (11.7–8.3 ka) periods.
Our starting point was a list of `r nrow(flora)` taxa (@tbl-results-summary) comprising the identifiable species observed at at least two Late Epipalaeolithic/Pre-Pottery Neolithic sites, according to our previous study of the regional archaeobotanical data [@ArranzOtaeguiRoe2023].
This was based on dataset collated from three previously published regional archaeobotanical databases: ADEMNES [@ADEMNES], ORIGINS [@ORIGINS], and COMPAG [@COMPAG].
We did not attempt to distinguish between the source of the remains [cf. @WallaceEtAl2018];
archaeobotanical assemblages are subject to a variety of preservational and recovery biases, so by no means were all the species on our list consumed or even deliberately collected by people.
However, we assume that there presence at a site of human settlement at least implies that they were part of the wider ecosystem that supported habitation there.

@@

flora <- left_join(flora, archaeo_pred)
```

@@

<!-- TODO expand/rewrite -->
Our results for the Levant are consistent with the current understanding of this region as developing early intensive foraging economies [the Natufian culture, @cite] and as a centre of origin of agriculture [@cite].
Within the Levant, many species show moderate reductions in ranges over the Pleistocene/Holocence boundary, retreating from the Badia/transjordan region [e.g. @fig-palaeodist-barely-levant].

Loss of the Northern Mesopotamia–Anatolia region from the predicted ranges of crop progenitors is interesting in light of the 'golden triangle' hypothesis [@Lev-YadunEtAl2000; @KozlowskiAurenche2005; @AbboEtAl2010], which puts this region at the centre of the development of agriculture.
Multiple lines of archaeological evidence have emerged that point away from this hypothesis and towards a more geographically diverse origin [@Asouti2006; @cites], but it remains the area with some of the earliest clear evidence of domestication [@ZoharyEtAl2012; @KabukcuEtAl2021; @UlasEtAl2024].
Comparative genetics also points to the Northern Mesopotamia region as the centre of diversity of many crops [e.g. @HaasEtAl2019].
But since these studies are based on modern genomes, if the wild range of these plants has, as our modelling suggests, shifted since the Early Holocene, the apparent centre of diversity may have shifted with them.
Our reconstructions are consistent with the late arrival of intensive plant-based foraging economies in this region (cf. the Natufian of the Levant),
and more broadly there need not be a link between the core *wild* range of a plant and the core area of its domestication. 
A scenario in which cultivation emerged at the edges of the ranges of valuable plant resources—as a means of extend their natural niche—is also plausible.

The near-absence of the Zagros in any predicted ranges is also surprising, given mounting evidence that domestication took place just as early in the eastern *Mashriq* as it did in the west [@Braidwood; @GanjDarehGoats; @HillyFlanksStuff].
We consider that the most likely explanation for this is that our flora does not include the species that were most important to plant subsistence in the east.
Archaeobotanical data on Neolithic sites in the Zagros is limited (compared to the Levant in particular) due to a hiatus in field research there from the 1980s to early 2010s [@cite].
Recent research [@Amaia?] indicates ... <!-- TODO: weird plants -->

Cyprus and the Aegean are not conventionally considered part of the primary zone of domestication but rather amongst the first regions that acquired agriculture from West Asia.
Our analysis complicates this picture, as it indicates that the wild ranges of many crop progenitors included these regions.
Early examples of several domesticates are recorded at sites on Cyprus, Western Anatolia and Greece [@ArranzOtaeguiRoe2023],
and the Aegean region was probably connected to West Asia by a land bridge via Anatolia until the Early Holocene [@AksuHiscott2022].
Were these area part of the same broader 'interaction sphere' [@cite] that produce Neolithic agriculture in West Asia?

@@

Almost all the cereal and legume crop progenitors we modelled are predicted to have only been found in the Levant during the terminal Pleistocene and Early Holocene (see appendix).
Part of this may be do with the fact that both our initial flora and training occurrence dataset have a strong bias towards the southern Levant, but strikingly the modelled *current* ranges of these plants do tend to include Anatolia and the Zagros, so this cannot be the only factor.
One notable exception is the wild ancestor of chickpea (*Cicer reticulatum*), which is predicted to have a distribution centered on Northern Mesopotamia but encompassing much of the the 'hilly flanks' (except the southern Levant, @fig-palaeodist-chickpea-rye). 
Another is rye (*Secale cereale*), which is inferred to be primarily Western Anatolian [@fig-palaeodist-chickpea-rye].
This is perhaps relevant to rye's unusual domestication history, as a crop of West Asian origin that was apparently not first cultivated until much later than the 'founder crops', in Europe [@SchreiberEtAl2021].

@@

Wild barley (*Hordeum spontaneum*) its relatives show a contraction of their predicted ranges from the Pleistocene to the Holocene,
concurrent with it being brought into cultivation [@fig-palaeodist-barely-levant].
It also sees a marked decline in the archaeobotanical record from the Early PPNA/Early PPNB (where it was amongst the most common taxa) to the Late PPNB and Late Neolithic [@ArranzOtaeguiRoe2023].
Pistachio (*Pistacia atlantica*) shows similar trends, but it is less certain that this species was cultivated in the Neolithic.
Conversely, the various wild wheat species native to West Asia show almost no response to Pleistocene/Holocene climate change, even within the Levant, and in the archaeobotanical record wheat displays the opposite trend to barley and pistachio – becoming gradually more abundant through the course of the Neolithic and dominant by its end.
We hypothesise that climate-linked range shifts were a factor driving these changes in the apparent economic important of different crops:
the decreasing availability of barley and pistachio in the terminal Pleistocene may have prompted cultivation as a strategy to retain access to them,
though in the long run it made them less attractive as staple crops compared to the more resilient wheat.
However, these dynamics are not seen in the majority of crop progenitors.

@@

"All models are wrong" [@cite], but the performance of our models on independent test datasets give confidence in its predictions, which are plausible and consistent with broad-scale patterns in the archaeological record.

* Results w.r.t. to specific archaeologically-attested species not to promising, but this doesn't mean the models are useless!
* Discrepencies suggests several areas for further research and methodological development

Modelling a large number of species using machine learning, the substantial occurrence datasets available for the present day, and a hindcasting approach to past distributions also represents a significant advance in the methodology of palaeoecological niche modelling. 
This approach is enabled by the availability of high quality, global open datasets in ecology [@GBIF2025; @GBIFSecretariat2023], earth science [@SRTM], and climatology [@@KargerEtAl2017; @BrownEtAl2018].
Unfortunately in relation to these fields open data in archaeology lags conspiciously behind.
Though we have benefitted from the relatively good state of publication of archaeobotanical compilations in our region of study [@ADEMNES; @COMPAG; @ORIGINS], 
further development of open, comprehensive and up-to-date 'backbone' datasets on site locations and chronologies is needed to advance archaeoecological modelling to the same level.

>>>>

---
title: Biogeography of crop progenitors and wild plant resources in the terminal Pleistocene and Early Holocene of West Asia, 14.7–8.3 ka
date: today
author: 
- name: Joe Roe
  id: jr
  orcid: 0000-0002-1011-1244
  email: joeroe@hey.com
  url: https://joeroe.io
  affiliations:
    - name: University of Bern
      country: Switzerland
    - name: University of Copenhagen
      country: Denmark
  corresponding: true
- name: Amaia Arranz-Otaegui
  id: aao
  orcid: 0000-0002-5091-6426
  affiliation: 
    - name: University of the Basque Country
      country: Spain
abstract: |
  This paper presents the first continuous, spatially-explicit reconstructions 
  of the palaeodistributions of 82<!-- TODO update --> plant species found regularly in association 
  with early agricultural archaeological sites in West Asia, including the 
  progenitors of the first crops. We used machine learning to train an 
  ecological niche model of each species based on its present-day distribution 
  in relation to climate and environmental variables. Predictions of the 
  potential ranges of these species at key stages of the Pleistocene–Holocene 
  transition could then be derived from these models using downsampled data from 
  palaeoclimate simulations.
  The models predict significant reductions and/or shifts in species ranges in
  the terminal Pleistocene and Early Holocene compared to present conditions. 
  Many species that are found throughout the region's 'hilly flanks' today 
  are indicated to have much more restricted distributions centered on the 
  Levant, Cyprus and Western Anatolia. In addition, overall ranges shrunk by an 
  average of c. 25% from the terminal Pleistocene to the Early Holocene. 
  Modelled ranges do not reliably predict the observed  occurrence of specific 
  species at archaeological sites, but are aligned with more general trends in
  the relative abundance of species.<!--
  however [...]
  The regional ubiquity of species in the archaeological record is [not] correlated with the 
  predicted size of its range and the diversity of archaeobotanical assemblages is [not] 
  correlated with the predicted diversity of its environs. This indicates that 
  trends in taxonomic composition of the archaeobotanical record is [not] likely 
  to have been influenced by environmental change and species turnover, in
  addition to human economic choices.-->
format: 
  elsevier-pdf:
    keep-tex: true
    # cite-method: citeproc
    journal:
      name: Open Quaternary
      formatting: preprint
      cite-style: authoryear
bibliography: references.bib
execute: 
  echo: false
---

@@

The Pleistocene–Holocene transition in West Asia marked a turning point in global environmental history, as humans brought the first plants under cultivation and began modifying surrounding ecosystems to support their own subsistence.
West Asia is part of the native range of a remarkable number of domesticable plant species, including wild relatives of wheat, barley, peas, lentils, and other crops of global importance [@Diamond2002].
These species supported uniquely dense and complex Late Epipalaeolithic (15–11.7 ka) societies based on intensive foraging [@BarYosef1998; @MaherEtAl2012] and eventually plant management and pre-domestication cultivation [@Harris2007].
As they were brought under domestication in the Pre-Pottery Neolithic period (11.7–8.5 ka), the world's first agriculture was shaped by the ecosystems from which it emerged and was embedded in.

Decades of research in archaeobotany and zooarchaeology have reconstructed the subsistence economies of Late Epipalaeolithic and Neolithic sites in great detail.
Together with studies of charcoal, pollen, soil, and a variety of palaeoclimate archives [@JonesEtAl2019], they also tell us much about the environments surrounding these settlements.
However, each of these sources of evidence is subject to the wide variety of taphonomic and recovery biases that are inherent in any direct record of the past.
They are also, by definition, records of the (human) environment at particular times and places.
Interpolating these snapshots to give a holistic picture of the regional ecologies is not straightforward – to date, it has tended to rely on non-explicit, inductive modelling.
The majority are also filtered through human action, producing a mixed single that makes it difficult to disentangle anthropic effects from the background of environmental change in this period of rapid climatic alteration.

@@

# Background
<!--
* The transition to agriculture in West Asia was...

@@

## Ecological niche modelling in archaeology
-->
Ecological niche modelling or species distribution modelling is widely used by ecologists to predict the geographic range of a species based on a set of environmental predictors [@FranklinMiller2009].
Essentially, it involves combining records of where an organism has been observed with environmental data (climate, topography, etc.) for those locations to model the range of environmental values at which that species  – its environmental niche.
This model can then be used to predict the range of the organism in question either in the same or a different environment.
@TownsendPetersonSoberon2012 suggests reserving the term 'species distribution modelling' for when the method is used to recover the verifiable range of a species in a real and existing environment, and using 'ecological niche modelling' as the broader term covering hypothetical or predictive applications – a convention we follow here when referring to predictive or 'hindcast' models of past ranges.
Within this overarching framework, ecological niche modelling encompasses a wide range of applications and a variety of potential environmental predictors, modelling approaches, and methodologies, which we will not attempt to review here.

@@

@FranklinEtAl2015 review palaeoecological niche modelling and advocate for its greater adoption in environmental archaeology.
In an early application to West Asia, @ConollyEtAl2012 used the occurrence of wild and domestic *Bos* remains at prehistoric archaeological sites to map the evolving niche of cattle over the Pleistocene–Holocene transition.
It has been used to model the availability of fauna exploited by humans at wider scales [e.g. @deAndresHerreroEtAl2018; @YaworskyEtAl2023] and, in a West Asian context, of foraged plant resources in the landscape around the Neolithic sites on the Konya Plain [@CollinsEtAl2018].
Modelling the spread of crops has been another significant archaeological application [e.g. @KrzyzanskaEtAl2022; @Krzyzanska2023], though not as yet applied to West Asia.

@@

The major practical limitation of the hindcasting approach is that it relies on spatially explicit, high resolution palaeoenvironmental surfaces with continuous coverage of the region and periods of interest.
Until recently, this has not been widely available for most applications, which is perhaps why only a minority of studies use it [cf. @KrzyzanskaEtAl2022; @YaworskyEtAl2023].
In this study, we are able to take advantage of the increasing availability of high resolution, global palaeoclimate data derived from simulation experiments with general circulation models of climate [@BrownEtAl2018; @BrownEtAl2020; @KargerEtAl2023].

@@

The aim of our study was to model the biogeography of species relevant to human subsistence economies in West Asia (excluding the Southern Arabian peninsula, see @fig-region) during the archaeological Late Epipalaeolithic (15–11.7 ka) and Pre-Pottery Neolithic (11.7–8.3 ka) periods.
Our starting point was a list of `r nrow(flora)` taxa (@tbl-results-summary) comprising the identifiable species observed at at least two Late Epipalaeolithic/Pre-Pottery Neolithic sites, according to our previous study of the regional archaeobotanical data [@ArranzOtaeguiRoe2023].
This was based on dataset collated from three previously published regional archaeobotanical databases: ADEMNES [@ADEMNES], ORIGINS [@ORIGINS], and COMPAG [@LucasFuller2018; @FullerEtAl2018; based on @ColledgeEtAl2004; @ShennanConolly2007].
We did not attempt to distinguish between the source of the remains [cf. @WallaceEtAl2018];
archaeobotanical assemblages are subject to a variety of preservational and recovery biases, so by no means were all the species on our list consumed or even deliberately collected by people.
However, we assume that there presence at a site of human settlement at least implies that they were part of the wider ecosystem that supported habitation there.

@@

flora <- left_join(flora, archaeo_pred, join_by(taxon))
```

@@

<!-- TODO expand/rewrite -->
Our results for the Levant are consistent with the current understanding of this region as developing early intensive foraging economies [the Natufian culture, @BarYosef1998] and as a centre of origin of agriculture [@Zeder2011a].
Within the Levant, many species show moderate reductions in ranges over the Pleistocene/Holocence boundary, retreating from the Badia/transjordan region (e.g. barley, fig. @fig-palaeodist-barely-levant).

Loss of the Northern Mesopotamia–Anatolia region from the predicted ranges of crop progenitors is interesting in light of the 'golden triangle' hypothesis [@Lev-YadunEtAl2000; @KozlowskiAurenche2005; @AbboEtAl2010], which puts this region at the centre of the development of agriculture.
Multiple lines of archaeological evidence have emerged that point away from this hypothesis and towards a more geographically diverse origin [@Asouti2006; @FullerEtAl2011; @ArranzOtaeguiEtAl2016], but it remains the area with some of the earliest clear evidence of domestication [@ZoharyEtAl2012; @KabukcuEtAl2021; @UlasEtAl2024].
Comparative genetics also points to the Northern Mesopotamia region as the centre of diversity of many crops [e.g. @HaasEtAl2019].
But since these studies are based on modern genomes, if the wild range of these plants has, as our modelling suggests, shifted since the Early Holocene, the apparent centre of diversity may have shifted with them.
Our reconstructions are consistent with the late arrival of intensive plant-based foraging economies in this region (cf. the Natufian of the Levant),
and more broadly there need not be a link between the core *wild* range of a plant and the core area of its domestication. 
A scenario in which cultivation emerged at the edges of the ranges of valuable plant resources—as a means of extend their natural niche—is also plausible.

The near-absence of the Zagros in any predicted ranges is also surprising, given mounting evidence that domestication took place just as early in the eastern *Mashriq* as it did in the west [@Zeder2024].
We consider that the most likely explanation for this is that our flora does not include the species that were most important to plant subsistence in the east.
Archaeobotanical data on Neolithic sites in the Zagros is limited (compared to the Levant in particular) due to a hiatus in field research there from the 1980s to early 2010s [@Zeder2024].
Recent research [@WhitlamEtAl2018; @GonzalezCarreteroEtAl2023] indicates that plant subsistence in this region was based on a distinct set of species, compared to the Levant and Anatolia.

Cyprus and the Aegean are not conventionally considered part of the primary zone of domestication but rather amongst the first regions that acquired agriculture from West Asia.
Our analysis complicates this picture, as it indicates that the wild ranges of many crop progenitors included these regions.
Early examples of several domesticates are recorded at sites on Cyprus, Western Anatolia and Greece [@ArranzOtaeguiRoe2023],
and the Aegean region was probably connected to West Asia by a land bridge via Anatolia until the Early Holocene [@AksuHiscott2022].
Were these area part of the same broader 'interaction sphere' that produce Neolithic agriculture in West Asia?

@@

Almost all the cereal and legume crop progenitors we modelled are predicted to have only been found in the Levant during the terminal Pleistocene and Early Holocene (see appendix).
Part of this may be do with the fact that both our initial flora and training occurrence dataset have a strong bias towards the southern Levant, but strikingly the modelled *current* ranges of these plants do tend to include Anatolia and the Zagros, so this cannot be the only factor.
One notable exception is the wild ancestor of chickpea (*Cicer reticulatum*), which is predicted to have a distribution centered on Northern Mesopotamia but encompassing much of the the 'hilly flanks' (except the southern Levant, @fig-palaeodist-chickpea-rye). 
Another is rye (*Secale cereale*), which is inferred to be primarily Western Anatolian (@fig-palaeodist-chickpea-rye).
This is perhaps relevant to rye's unusual domestication history, as a crop of West Asian origin that was apparently not first cultivated until much later than the 'founder crops', in Europe [@SchreiberEtAl2021].

@@

Wild barley (*Hordeum spontaneum*) its relatives show a contraction of their predicted ranges from the Pleistocene to the Holocene,
concurrent with it being brought into cultivation (@fig-palaeodist-barely-levant).
It also sees a marked decline in the archaeobotanical record from the Early PPNA/Early PPNB (where it was amongst the most common taxa) to the Late PPNB and Late Neolithic [@ArranzOtaeguiRoe2023].
Pistachio (*Pistacia atlantica*) shows similar trends, but it is less certain that this species was cultivated in the Neolithic.
Conversely, the various wild wheat species native to West Asia show almost no response to Pleistocene/Holocene climate change, even within the Levant, and in the archaeobotanical record wheat displays the opposite trend to barley and pistachio – becoming gradually more abundant through the course of the Neolithic and dominant by its end.
We hypothesise that climate-linked range shifts were a factor driving these changes in the apparent economic important of different crops:
the decreasing availability of barley and pistachio in the terminal Pleistocene may have prompted cultivation as a strategy to retain access to them,
though in the long run it made them less attractive as staple crops compared to the more resilient wheat.
However, these dynamics are not seen in the majority of crop progenitors.

@@

"All models are wrong" [@Box1976], but the performance of our models on independent test datasets give confidence in its predictions, which are plausible and consistent with broad-scale patterns in the archaeological record.
<!--
* Results w.r.t. to specific archaeologically-attested species not to promising, but this doesn't mean the models are useless!
* Discrepencies suggests several areas for further research and methodological development
 --> 
Modelling a large number of species using machine learning, the substantial occurrence datasets available for the present day, and a hindcasting approach to past distributions also represents a significant advance in the methodology of palaeoecological niche modelling. 
This approach is enabled by the availability of high quality, global open datasets in ecology [@GBIF2025; @GBIFSecretariat2023], earth science [@SRTM], and climatology [@KargerEtAl2017; @BrownEtAl2018].
Unfortunately in relation to these fields open data in archaeology lags conspiciously behind.
Though we have benefitted from the relatively long tradition of compiling archaeobotanical data in our region of study [@ColledgeEtAl2004; @ShennanConolly2007; @ADEMNES; @LucasFuller2018; @FullerEtAl2018; @WallaceEtAl2018; @ORIGINS], 
further development of open, comprehensive and up-to-date 'backbone' datasets on site locations and chronologies is needed to advance archaeoecological modelling to the same level.
